An Algorithmic Introduction to Lagrangian Coherent Structures

### 2014-01-24 main.tex
reorder theory section, add to ocean example

We find coherent Lagrangian vortices by considering strain, as was done in~\textcite{haller13:_coher_lagran}. To summarise the results in that publication, we examine closed curves $\gamma$ and $O(\epsilon)$ neighbourhoods or ``belts'' around them. For most $\gamma$, neighbouring belts of $O(\epsilon)$ exhibit strain variability of $O(\epsilon)$. Curves that form coherent Lagrangian vortices are exceptional however and display no observable variability in their average strain within $O(\epsilon)$ neighbourhoods.

The mathematical identification of a coherent Lagrangian vortex para\-metrises a curve $\gamma$ by $\boldsymbol r(s)$ with $s \in [0,\sigma]$ at time $t_0$. The length of a tangent vector $\boldsymbol r'(s)$ is $l_{t_0}(s)$. The advected curve $F_{t_0}^t(\gamma)$ has a tangent vector $(\text{d}/\text{d}s) F_{t_0}^t(\boldsymbol r(s))$ and a length we denote by $l_t(s)$. The two tangent lengths are calculated:
\begin{gather*}
l_{t_0} = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
l_t = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
\end{gather*}
where $\langle \cdot, \cdot \rangle$ denotes the Euclidean inner product.
The average tangential strain along $\gamma$ is:
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s.
\]

Coherent Lagrangian vortices are associated with the strain tensor
\[
E_\lambda(\boldsymbol x_0) = \frac12 [C_{t_0}^t(\boldsymbol x_0) - \lambda^2 I]
\]
and curves uniformly stretched by the same factor $\lambda$ when advected by the flow from time $t_0$ to $t$ are closed trajectories of one of two explicit 
differential equations: 
\begin{equation}
\boldsymbol r' = \boldsymbol \eta_\pm(\boldsymbol r),
\label{eq:etafields}
\end{equation}
where
\begin{equation}
\boldsymbol \eta_\pm = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \boldsymbol \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 
- \lambda_1}} \boldsymbol \xi_2.
\label{eq:eta}
\end{equation}
We call the curves associated with coherent Lagrangian vortices lambda-lines.

Hyperbolic LCSs are given by a variational approach based on shear presented in~\textcite{farazmand13:_shearless}. The Lagrangian shear is calculated as:
\[
p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
\]
where
\[
D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
The average Lagrangian shear along a curve $\gamma$ over the time interval $[t_0,t]$ is:
\[
P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s.
\]

Similarly to Lagrangian vortices, exceptional hyperbolic curves without shear in $O(\epsilon)$ neighbourhoods are null-geodesics of the Lorentzian metric
\[
g(\boldsymbol u, \boldsymbol v) = \langle \boldsymbol u, D(\boldsymbol x_0) \boldsymbol v \rangle.
\]
Repelling LCSs are obtained as solutions of the strainline differential equation:
\[
\boldsymbol r' = \boldsymbol \xi_1(\boldsymbol r),
\]
and attracting LCSs are obtained as a solution of the stretchline differential equation:
\[
\boldsymbol r' = \boldsymbol \xi_2(\boldsymbol r).
\]

@@

This section presents the numerical methods used to compute Lagrangian coherent structures. In other words, the algorithms that constitute LCS Tool are presented.

@@

To obtain Lagrangian coherent structures the Cauchy-Green strain tensor must be calculated first. More specifically, the eigenvalues and eigenvectors of the Cauchy-Green strain tensor are needed. The function associated with this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps of this function are enumerated in Table~\ref{t:Cauchy-Green algorithm} and Table~\ref{t:eig_cgStrain syntax} summarises the function syntax.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\\
\hline
\lstinline!derivative! & function handle for flow velocity equations\\
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\\
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\\
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\\
\hline
\lstinline!incompressibile! & optional logical to specify if incompressibility is imposed. Default is  \lstinline!false!.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration options\\
\hline \hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

We setup a Cartesian grid according to user-defined resolutions. We can find the optimal resolution by an iterative procedure in which we start with a low resolution and double it until convergence of LCSs is observed visually. A resolution of about 500 grid point along the longest axis usually gives good results. Furthermore an auxiliary grid is setup. This is illustrated in Figure~\ref{f:main and auxiliary grids}. The improvements the auxiliary grid yields compared to using only the main grid were reported in \textcite{farazmand12:_comput_lagran}. Experience indicates that optimal results are obtained when the auxiliary grid spacing is between 1 and 10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of main and auxiliary grid used to compute the Cauchy-Green strain. Solid circles are main grid points and hollow circles are auxiliary grid points.}
\label{f:main and auxiliary grids}
\end{figure}

@@

As stated, typically we set the main grid resolution of the Cauchy-Green strain tensor to around 500 grid points. For a square domain, this implies the velocity equations must be integrated, after including 4 auxiliary grid points around every main grid point, at 1.25 million grid points. The ``natural'' way to perform these integration calculations would be to solve a two dimensional first-order ODE system at every grid point and iterate over every grid point within a for-loop. With 1.25 million grid points this calculation can take an excessively long time. To alleviate this problem, we solve the flow equations in a vector form where the equations at all grid points are combined into a single system of equation, even though the system of composed of independent blocks of two dimensional first-order ODE systems. MATLAB's \lstinline!ode45! function is used to perform the integration and the flow integration typically takes five to ten minutes with default error tolerances. The drawback of vector form integration is its memory requirements can become excessive at high resolutions, furthermore, writing the velocity function in vector form is more error-prone than the intrinsic two dimensional form.

The Cauchy-Green strain tensor can have singularities (i.e. $C_{t_0}^t(\boldsymbol x_0) = I$). Such points are isolated in general\parencite{delmarcelle94}, therefore no special accommodations for are made in computations for singularities of the Cauchy-Green strain tensor.

@@

At some grid points it may occur that $\lambda_2 < 1$ due to numerical integration errors of the flow velocity equations. By setting integration tolerances to smaller values the number of grid points with $\lambda_2 < 1$ can be reduced. Nonetheless, some grid points can be so close to singularities of the Cauchy-Green strain tensor that satisfying $\lambda_2 >= 1$ everywhere could be excessively computationally costly. Therefore the function \lstinline!eig_cgStrain! records the number of points at which incompressibility cannot be imposed and this measure can also help to set appropriate integration tolerances.

@@

Velocity fields defined by datasets require pre-processing prior to performing the integration that yields the flow map. An interpolation function must be written to enable evaluating the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. MATLAB's interpolation functions can be integrated into the velocity function. As with analytic cases, care should be taken to ensure the interpolation embedded in the flow velocity integration function works correctly when performing integration in vector form.

@@

Shear LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. Since $\lambda = 1$, the closed curves are non-stretching and area preserving. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the first return map. A lambda-line closing onto itself in the starting point is a shear LCS.

The main steps of the algorithm used to calculate shear LCS are enumerated in Table~\ref{t:Shear LCS algorithm}. The syntax of shear LCS functions in LCS Tool is in Table~\ref{t:Shear LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain
\item Calculate Poincare return map
\item Find closed orbits in Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate Shear LCS}
\label{t:Shear LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional positive integer. Default is 5.\\
\hline
\lstinline!dThresh! & optional scalar. Default is $10^{-2}$.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration options\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool shear LCS functions syntax}
\label{t:Shear LCS functions}
\end{table}

@@

\begin{table}
\begin{enumerate}
\item Linearly interpolate vector field orientation at initial position
\item At next position, check whether vector field has rotated by over 90\degree, if yes, flip the vector field orientation by 180\degree.
\item Stop integration when lambda-line returns to Poincare section, or maximum integration length has been reached.
\end{enumerate}
\caption{Algorithm used for variable time step integration of lambda-lines.}
\label{t:variable step integration}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{graphics/variable_step_integration}
\end{center}
\caption{Schematic illustration of lambda-line variable time step integration. At the initial point, there is an orientation discontinuity at the lower right grid point that must be corrected prior to linear interpolation. At point \#1, no orientation discontinuties are present. At point \#2 the interpolated eigenvector field must be rotated by 180\degree\,to match the orientation at the previous point, \#1.}
\label{f:variable step integration}
\end{figure}

An example of a Poincare return map produced from integration of the $\boldsymbol \eta_\pm$ field is shown in Figure~\ref{f:Poincare return map}. Well behaved orbits will return to the Poincare section and their integration will be stopped using an ODE event detection function. Some orbits may however deviate far from the Poincare section and their integration could take inordinately long. To prevent this behaviour, a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!, is specified. In practice, viewing the Poincare section as the radius of a circle and setting the maximum lambda-line integration length to twice the circumference gives acceptable results. In Figure~\ref{f:Poincare return map}, circle markers indicate closed orbit positions. Closed orbits exist where the Poincare return map is zero. The function \lstinline!poincare_closed_orbit! performs the computations. In Figure~\ref{f:Poincare return map} not all zero crossings have circles. This is because a threshold parameter, \lstinline!dThresh! whose default value is 1\% of the length of the Poincare section, is used to discard those zero crossings that appear in regions with high numerical noise. The position of each detected zero crossing is refined by iteratively bisecting the interval of the zero crossing. If after a predetermined number of iterations, \lstinline!nBisection! whose default value is five, the two points around the zero crossing have absolute values of more \lstinline!dThresh!, the zero crossing is discarded. In the Figure, discarded zero crossings are seen where $0 < s < 0.05$. Once all closed lambda-line orbits have been located, the outermost orbit of every Poincare section is deemed a shear LCS.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare return map obtained for shear LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost closed orbit and shear LCS.}
\label{f:Poincare return map}
\end{figure}

@@

@@

\section{Examples}

\subsection{The double gyre}

The double gyre is a model for a pattern observed in geophysical flows\parencite{shadden05:_defin_lagran_lyapun}. The velocity equations are:
\begin{equation}
\begin{split}
\dot x = -\pi A \sin[\pi f(x,t)] \cos(\pi y)\\
\dot y = \pi A \cos[\pi f(x,t)] \sin(\pi y) \frac{\partial f(x,t)}{\partial x}\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function corresponding to these equations is given in Listing~\ref{l:double gyre derivative} and shows how to specify a derivative function that supports vector form integration.

>>>>

\subsection{Elliptic LCS}
We find coherent Lagrangian vortices as closed orbits of one of the two vector fields
\begin{equation}
\boldsymbol \eta^{\lambda}_\pm = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \boldsymbol \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 
- \lambda_1}} \boldsymbol \xi_2,
\label{eq:eta}
\end{equation}
which are a function of the eigenvalues and eigenvectors of $C_{t_0}^t(\boldsymbol x_0)$.
Curves $\gamma$ that are trajectories of $\boldsymbol \eta^{\lambda}_\pm$ can be parametrized as $\boldsymbol r(s)$ with their arclength $s \in [0,\sigma]$. They are given by integrating
\begin{equation}
\boldsymbol r' \equiv \frac{d \boldsymbol r}{ds} = \boldsymbol \eta^{\lambda}_\pm(\boldsymbol r).
\label{eq:etafields}
\end{equation}

$\lambda$ is a free parameter close to unity that indicates the uniform stretching of the curve $\gamma$, i.e., each tangent vector $\boldsymbol r'(s)$ with length $l_{t_0}$ stretches by a factor $\lambda$ to length $l_{t}$ under the flow map:
\begin{gather*}
l_{t_0}(s) = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
l_t(s) = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
l_t(s) = \lambda\, l_{t_0}(s).
\end{gather*}
For $\lambda=1$, $\boldsymbol \eta^{\lambda}_\pm$ points in the direction of maximal Lagrangian shear and we obtain uniformly non-stretching curves over the finite time interval $[t_0, t]$. 
In summary, the boundaries of coherent Lagrangian vortices are closed curves that uniformly stretch by a factor $\lambda$ close to unity when advected by the flow. We call them lambda-lines. The outermost lambda-line is the physically observable boundary of an eddy, generating the tracer pattern of an isolated rotating fluid region.

Note that $\boldsymbol \eta^{\lambda}_\pm$ follows from the variational principle that the first variation of the averaged Lagrangian strain along the searched closed curve vanishes (see \parencite{haller13:_coher_lagran} for full derivations):
\begin{eqnarray}
Q(\gamma) &=& \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s \\
\delta Q(\gamma) &=& 0.
\end{eqnarray}
This coherence principle means that the searched curve $\gamma$, the coherent eddy boundary, is a stationary point of the averaged Lagrangian strain functional and therefore nearby fluid elements cannot break away from the curve $\gamma$.

%We find coherent Lagrangian vortices by considering strain, as was done in~\textcite{haller13:_coher_lagran}. To summarise the results in that publication, we examine closed curves $\gamma$ and $O(\epsilon)$ neighbourhoods or ``belts'' around them. For most $\gamma$, neighbouring belts of $O(\epsilon)$ exhibit strain variability of $O(\epsilon)$. Curves that form coherent Lagrangian vortices are exceptional however and display no observable variability in their average strain within $O(\epsilon)$ neighbourhoods.
%
%The mathematical identification of a coherent Lagrangian vortex para\-metrises a curve $\gamma$ by $\boldsymbol r(s)$ with $s \in [0,\sigma]$ at time $t_0$. The length of a tangent vector $\boldsymbol r'(s)$ is $l_{t_0}(s)$. The advected curve $F_{t_0}^t(\gamma)$ has a tangent vector $(\text{d}/\text{d}s) F_{t_0}^t(\boldsymbol r(s))$ and a length we denote by $l_t(s)$. The two tangent lengths are calculated:
%\begin{gather*}
%l_{t_0} = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
%l_t = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
%\end{gather*}
%where $\langle \cdot, \cdot \rangle$ denotes the Euclidean inner product.
%The average tangential strain along $\gamma$ is:
%\[
%Q(\gamma) = \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s.
%\]
%
%Coherent Lagrangian vortices are associated with the strain tensor
%\[
%E_\lambda(\boldsymbol x_0) = \frac12 [C_{t_0}^t(\boldsymbol x_0) - \lambda^2 I]
%\]
%and curves uniformly stretched by the same factor $\lambda$ when advected by the flow from time $t_0$ to $t$ are closed trajectories of one of two explicit 
%differential equations: 
%\begin{equation}
%\boldsymbol r' = \boldsymbol \eta_\pm(\boldsymbol r),
%\label{eq:etafields}
%\end{equation}
%where
%\begin{equation}
%\boldsymbol \eta_\pm = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \boldsymbol \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 
%- \lambda_1}} \boldsymbol \xi_2.
%\label{eq:eta}
%\end{equation}
%We call the curves associated with coherent Lagrangian vortices lambda-lines.

\subsection{Hyperbolic LCS}
Hyperbolic LCS are trajectories of the two eigenvector fields $\boldsymbol \xi_{1,2}$ of the Cauchy-Green strain tensor $C_{t_0}^t$. Repelling hyperbolic LCSs, denoted as strainlines, are obtained as solutions of the differential equation
\[
\boldsymbol r_1' = \boldsymbol \xi_1(\boldsymbol r),
\]
and attracting hyperbolic LCSs, denoted as stretchlines, are obtained as a solution of the differential equation
\[
\boldsymbol r_2' = \boldsymbol \xi_2(\boldsymbol r).
\]
A normal perturbation to the strainline segment $\boldsymbol r_1'$ grows under the flow map by a factor $\lambda_2$, i.e., fluid elements are repelled by that rate in normal direction. Equivalently, a normal perturbation to the stretchline segment $\boldsymbol r_2'$ shrinks with factor $\lambda_1$ and fluid elements are attracted.

Similarly to ellitpic LCS, hyperbolic LCS are also based on a variational principle, in this case for the Lagrangian shear
\[
p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
\]
where
\[
D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
The first variation of the average Lagrangian shear along a hyperbolic LCS vanishes
\[
P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s.
\]
\[
\delta P(\gamma) = 0.
\]
They are one type of shearless LCS. For details of the derivation of hyperbolic LCS, see \cite{farazmand13:_shearless}.

%Hyperbolic LCSs are given by a variational approach based on shear presented in~\textcite{farazmand13:_shearless}. The Lagrangian shear is calculated as:
%\[
%p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
%\]
%where
%\[
%D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
%\]
%The average Lagrangian shear along a curve $\gamma$ over the time interval $[t_0,t]$ is:
%\[
%P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s.
%\]
%
%Similarly to Lagrangian vortices, exceptional hyperbolic curves without shear in $O(\epsilon)$ neighbourhoods are null-geodesics of the Lorentzian metric
%\[
%g(\boldsymbol u, \boldsymbol v) = \langle \boldsymbol u, D(\boldsymbol x_0) \boldsymbol v \rangle.
%\]
%Repelling LCSs are obtained as solutions of the strainline differential equation:
%\[
%\boldsymbol r' = \boldsymbol \xi_1(\boldsymbol r),
%\]
%and attracting LCSs are obtained as a solution of the stretchline differential equation:
%\[
%\boldsymbol r' = \boldsymbol \xi_2(\boldsymbol r).
%\]

@@

This section presents the numerical methods used to compute LCS and explains them step by step. In other words, the algorithms that constitute LCS Tool are presented.

@@

To obtain Lagrangian coherent structures the Cauchy-Green strain tensor must be calculated first. More specifically, the eigenvalues and eigenvectors of the Cauchy-Green strain tensor are needed. The function associated with this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps of this function are enumerated in Table~\ref{t:Cauchy-Green algorithm} and are described in detail in the following part. Table~\ref{t:eig_cgStrain syntax} summarises the function syntax.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\\
\hline
\lstinline!derivative! & function handle for flow velocity equations\\
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\\
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\\
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\\
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default is  \lstinline!false!.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration options\\
\hline \hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

We setup a Cartesian grid according to user-defined resolutions. We can find the optimal resolution by an iterative procedure in which we start with a low resolution and double it until convergence of LCSs is observed visually. 
If the chosen domain only comprises few expected LCS, e.g., one vortex, a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, for larger domains, a higher resolution must be chosen. Furthermore an auxiliary grid is setup. This is illustrated in Figure~\ref{f:main and auxiliary grids}. The improvements the auxiliary grid yields compared to using only the main grid were reported in \textcite{farazmand12:_comput_lagran}. Experience indicates that optimal results are obtained when the auxiliary grid spacing is between 1\% and 10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of main grid (filled circles) and auxiliary grid (empty circles) used to compute the Cauchy-Green strain. Solid circles are main grid points and hollow circles are auxiliary grid points. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

@@

As stated, typically we set the main grid resolution of the Cauchy-Green strain tensor to around 500 grid points. For a square domain, this implies the velocity equations must be integrated, after including 4 auxiliary grid points around every main grid point, at 1.25 million grid points. The ``natural'' way to perform these integration calculations would be to solve a two dimensional first-order ODE system at every grid point and iterate over every grid point within a for-loop. With 1.25 million grid points this calculation can take an excessively long time. To alleviate this problem, we solve the flow equations in a vector form where the equations at all grid points are combined into a single system of equations, even though the system is composed of independent blocks of two dimensional first-order ODE systems. MATLAB's \lstinline!ode45! function is used to perform the integration and the flow integration typically takes five to ten minutes with default error tolerances. The drawback of vector form integration is its memory requirements can become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the intrinsic two dimensional form.

The Cauchy-Green strain tensor can have singularities, i.e., $C_{t_0}^t(\boldsymbol x_0) = I$. Such points are isolated in general\parencite{delmarcelle94}, therefore no special accommodations for are made in computations for singularities of the Cauchy-Green strain tensor.

@@

At some grid points it may occur that $\lambda_2 < 1$ due to numerical integration errors of the flow velocity equations. By setting integration tolerances to smaller values the number of grid points with $\lambda_2 < 1$ can be reduced. Nonetheless, some grid points can be so close to singularities of the Cauchy-Green strain tensor that satisfying $\lambda_2 \geq 1$ everywhere could be excessively computationally costly. Therefore the function \lstinline!eig_cgStrain! records the number of points at which incompressibility cannot be imposed and this measure can also help to set appropriate integration tolerances.

@@

Velocity fields defined by datasets require pre-processing prior to performing the integration that yields the flow map. An interpolation function must be written to enable evaluating the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. MATLAB's interpolation functions can be integrated into the velocity function. As with analytic cases, care should be taken to ensure the interpolation embedded in the flow velocity integration function works correctly when performing integration in vector form. In section \ref{sec:oceandataset}, we present an example how LCS are computed for an ocean dataset and details of the interpolation function are given.

@@

Shear LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. Since $\lambda = 1$, the closed curves are non-stretching and area preserving. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the first return map. A lambda-line closing onto itself in the starting point is a shear LCS.
The main steps of the algorithm used to calculate shear LCS are enumerated in Table~\ref{t:Shear LCS algorithm} and described in detail in the following part. The syntax of shear LCS functions in LCS Tool can be seen in Table~\ref{t:Shear LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain as initial positions of lambda-lines
\item Integrate lambda-lines tangent to $\boldsymbol \eta_{\pm}$
\item Calculate Poincare return map
\item Find closed orbits in Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate Shear LCS}
\label{t:Shear LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional positive integer. Default is 5. Number of bisection steps to refine zero crossings of Poincare return map.\\
\hline
\lstinline!dThresh! & optional scalar. Default is $10^{-2}$. Threshold to discard discontinuous zero crossings of Poincare return map.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration options\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool shear LCS functions syntax}
\label{t:Shear LCS functions}
\end{table}

@@

\begin{table}
\begin{enumerate}
\item Linearly interpolate vector field orientation at initial position
\item At next position, check whether vector field has rotated by over 90\degree, if yes, flip the vector field orientation by 180\degree.
\item Stop integration when lambda-line returns to Poincare section, lambda-line hits the domain boundary, or maximum integration length has been reached.
\end{enumerate}
\caption{Algorithm used for variable time step integration of lambda-lines.}
\label{t:variable step integration}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/variable_step_integration}
\end{center}
\caption{Schematic illustration of lambda-line variable time step integration. At the initial point, there is an orientation discontinuity at the lower right grid point that must be corrected prior to linear interpolation. At point \#1, no orientation discontinuties are present. At point \#2 the interpolated $\eta_\pm$ vector must be rotated by 180\degree\,to match the orientation of the trajectory.}
\label{f:variable step integration}
\end{figure}

An example of a Poincare return map produced from integration of the $\boldsymbol \eta_\pm$ field is shown in Figure~\ref{f:Poincare return map}. Well behaved orbits will return to the Poincare section and their integration will be stopped using an ODE event detection function. Some orbits may however deviate far from the Poincare section and their integration could take inordinately long. To prevent this behaviour, a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!, is specified. In practice, viewing the Poincare section as the radius of a circle and setting the maximum lambda-line integration length to twice the circumference gives acceptable results. In Figure~\ref{f:Poincare return map}, circle markers indicate closed orbit positions. Closed orbits exist where the Poincare return map, the distance between the final and the initial point of the orbit $P(s)-s$, is zero. The function \lstinline!poincare_closed_orbit! performs the computations. In Figure~\ref{f:Poincare return map} not all zero crossings have circles. 
This is because a threshold parameter \lstinline!dThresh! is used to discard those zero crossings that appear in regions with high numerical noise. 
Its default value is 1\% of the length of the Poincare section.
The position of each detected zero crossing is refined by iteratively bisecting the interval of the zero crossing. If after a predetermined number of iterations, \lstinline!nBisection! whose default value is five, the two points around the zero crossing have absolute values of more \lstinline!dThresh!, the zero crossing is discarded. In Figure~\ref{f:Poincare return map}, discarded zero crossings are seen where $0 < s < 0.05$. Once all valid closed lambda-line orbits have been located, the outermost orbit of every Poincare section is deemed a shear LCS.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare return map obtained for shear LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost closed orbit and shear LCS.}
\label{f:Poincare return map}
\end{figure}

@@

\paragraph{Hyperbolic LCS inside ellitpic LCS}
Explain, figure?

@@

\section{Example}
In this section we present three examples of how to use LCS-Tool to obtain LCSs in different flows, namley the double gyre flow, the Bickley jet flow, and an oceanic geostrophic flow. All three examples are available as demo files in the demo folder of LCS-Tool and the user can follow the computation of LCS step by step.
